Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-12 ThesisJP/Background/background.tex
Started on comments from Ines on the introduction section.

\section{Volume rendering}\label{sec:volren}
\begin{figure}[t!]
\centering
\subfloat[Ray-casting]{
	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
}
\subfloat[Opacity transfer function]{
	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
}
\caption{The principle behind volume ray casting. a) A ray is casted out from every pixel in the output image. Along the ray, samples are gathered from the volume. The samples are assigned a color and an opacity before they are combined into one pixel value. b) The opacity transfer function used to determine how opaque a sample should be.}
\label{fig:vr}
\end{figure}
A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user. One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers. Another way is to project the whole volume into a two-dimensional image. The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image. Many different methods for volume rendering exist \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.

@@

The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is \cite{rtvg2006}
\begin{align}
I(s_0, s_{i+1}) &= I(s_0, s_{i}) + (1 - \alpha_{\text{acc},i}) c_{i+1}\\
\alpha_{\text{acc},i+1} &= \alpha_{\text{acc},i} + (1 - \alpha_{\text{acc},i}) \alpha(c_i+1).
\label{eq:fronToBack}
\end{align}
Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color intensity of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined. If $\alpha(v)$ is close to one for one or two consecutive samples, $\alpha_{\text{acc}}$ will quickly approach a value of one. An $\alpha_{acc,k}$-value close to one means that the contribution from $c_{k+1}, \cdots, c_{N_s-1}$ to the final value, $I(s_0, s_{N_s-1})$, will be minimal. Hence, the iteration can stop at $i=k$ without loosing visible samples. This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up our rendering algorithm if an opaque material is rendered.

@@

The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}. All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator. The Field II simulator applies a far-field approximation where each element is divided into mathematical elements. All field points needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse. For $w=\lambda/2$ the approximation is valid for $l>\lambda/16$.}. It is then possible to approximate the spatial impulse response for a given probe by summing a trapezoid impulse response function originating from each mathematical element in all field points \cite{Jensen1992}. A rough estimate of the number of required FLOPS for calculating the spatial impulse respons as it is done in Field II is:
\begin{align}
\text{FLOPS}_\text{Field II} = f_s\frac{w}{c}N_{\text{elements}}N_{\text{field points}},
\end{align}
where $f_s$ is the sampling frequency, $w$ is the width of the mathematical elements, $f_s\frac{w}{c}$ is the maximal number of samples in each trapezoid function, $N_{\text{elements}}$ is the total number of mathematical elements, and $N_{\text{field points}}$ is the number of field points. The required number of FLOPS for a simulation at $100$ MHz with $500$, $\lambda/2$ wide, mathematical elements across a field of $0.5$ million points is $6$ GFLOPS in order for it to complete after one second. If we want to simulate an ultrasound image the spatial impulse response on both transmit and recieve have to be calculated and convolved in time. The respons will also be different for each image line. Hence, more than 800 GFLOPS is required in order to complete a Field II simulation with the given example values of an image consisting of 70 image lines in one second. %However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}. 
Other simulators have also been introduced which are faster but less exact than Field II \cite{Hergum2009}.

@@

The above examples show that realistic and dense ultrasound simulations are computationally demanding. On the other hand we find that the required amount of processing is within the theoretical limits of modern CPUs and GPUs. Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions. In addition to \textbf{Paper V} there are several simulators that have been presented the last years, which does exactly this by leveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.

\section{Concluding remarks}
All the presented problems in this chapter have in common that they require a processing throughput from ten to several hundred GFLOPS. Comparing this to the typical clock frequency of modern CPUs and GPUs  at 3 and 1.5 GHz we see that GFLOPS problems require parallel execution in order to achieve real-time performance. With CPU threads it is only possible to achieve 24 GFLOPS in peak performance on a eight-core CPU. To reach a peak performance above 100 GFLOPS one has to use SIMD instructions on a CPU or SIMT instructions on a GPU. It also must be possible to divide the problem into a large number of independent calculations.

>>>>

\section{Volume rendering}\label{sec:volren}
\begin{figure}[t!]
\centering
\subfloat[Ray-casting]{
	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
}
\subfloat[Opacity transfer function]{
	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
}
\caption{The principle behind volume ray casting. a) A ray is casted out from every pixel in the output image. Along the ray, samples are gathered from the volume. The samples are assigned a color and an opacity before they are combined into one pixel value. b) The opacity transfer function used to determine how opaque a sample should be.}
\label{fig:vr}
\end{figure}
A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user. One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers. Another way is to project the whole volume into a two-dimensional image. The latter technique is known as volume rendering, and it presents to the user the three dimensional structure of the heart in one two-dimensional image. Many different methods for volume rendering exist \cite{brodlie2001}, but in this thesis we will focus on ray casting \cite{Levoy1988}.

@@

The iterative composition scheme for front-to-back ray casting of one ray, from $s_0$ to $s_{N_s-1}$ where $s$ is the sample position and $N_s$ is the number of samples along the ray, is \cite{rtvg2006}
\begin{align}
I(s_0, s_{i+1}) &= I(s_0, s_{i}) + (1 - \alpha_{\text{acc},i}) c_{i+1}\\
\alpha_{\text{acc},i+1} &= \alpha_{\text{acc},i} + (1 - \alpha_{\text{acc},i}) \alpha(c_i+1).
\label{eq:fronToBack}
\end{align}
Here, $\alpha(v)$ is the opacity transfer function, $c_{k}$ is the color intensity of sample $s_{k}$, $\alpha_{\text{acc}}$ is the accumulated opacity, and $I(s_0, s_{N_s-1})$ is the intensity value of all the projected samples combined. If $\alpha(v)$ is close to one for one or two consecutive samples, $\alpha_{\text{acc}}$ will quickly approach a value of one. An $\alpha_{acc,k}$-value close to one means that the contribution from $c_{k+1}, \cdots, c_{N_s-1}$ to the final value, $I(s_0, s_{N_s-1})$, will be minimal. Hence, the iteration can stop at $i=k$ without loosing visible samples. This optimization technique is known as \textit{early-ray-termination}, and will significantly speed up the rendering algorithm if an opaque material is rendered.

@@

The most popular simulator for medical ultrasound field simulations is Field II \cite{Jensen1992}. All simulations in both \textbf{Paper II} and \textbf{III} are produced with this simulator. The Field II simulator applies a far-field approximation where each element is divided into mathematical elements. All field points then needs to be in the far-field of these small elements\footnote{$l\gg\frac{w^2}{4\lambda}$, where $w$ is the mathematical element width, $l$ is the distance to the field point, and $\lambda$ is the wavelength of the transmitted pulse. For $w=\lambda/2$ the approximation is valid for $l>\lambda/16$.}. It is then possible to approximate the spatial impulse response for a given probe by summing a trapezoid impulse response function originating from each mathematical element in all field points \cite{Jensen1992}. A rough estimate of the number of required FLOPS for calculating the spatial impulse response as it is done in Field II is:
\begin{align}
\text{FLOPS}_\text{Field II} = f_s\frac{w}{c}N_{\text{elements}}N_{\text{field points}},
\end{align}
where $f_s$ is the sampling frequency, $w$ is the width of the mathematical elements, $f_s\frac{w}{c}$ is the maximal number of samples in each trapezoid function, $N_{\text{elements}}$ is the total number of mathematical elements, and $N_{\text{field points}}$ is the number of field points. The required number of FLOPS for a simulation at $100$ MHz with $500$, $\lambda/2$ wide, mathematical elements across a field of $0.5$ million points is $6$ GFLOPS in order for it to complete after one second. If we want to simulate an ultrasound image the spatial impulse response on both transmit and receive have to be calculated and convolved in time. The response will also be different for each image line. Hence, more than 800 GFLOPS is required in order to complete a Field II simulation in one second with the given example values and an image consisting of 70 image lines. %However, Jensen and Nikolov have shown ways of reducing the computational load of Field II simulations \cite{Jensen2000}. 
Other simulators have also been introduced which are faster but less exact than Field II \cite{Hergum2009}.

@@

The above examples show that realistic and dense ultrasound simulations are computationally demanding. On the other hand we find that the required amount of processing is within the theoretical limits of modern CPUs and GPUs. Since ultrasound field simulations are quite easy to accelerate in parallel it should be possible to add increased interactivity to ultrasound simulations tools by utilizing parallel programing and SIMD instructions. In addition to \textbf{Paper V}, several simulators have recently accomplished this by leveraging the powers of GPUs \cite{Gjerald2012, Reichl2009, Hlawitschka2011}.

\section{Concluding remarks}
All the presented problems in this chapter, except DAS beamforming, have in common that they require a processing throughput of more than ten GFLOPS. Comparing this to the typical clock frequency of modern CPUs at 3 GHz we see that GFLOPS problems will require parallel execution in order to achieve real-time performance. With CPU threads the theoretical peak performance on an eight-core CPU increases to 24 GFLOPS. However, to reach a peak performance above 100 GFLOPS, SIMD instructions on a CPU or SIMT instructions on a GPU have to be used. It also must be possible to divide the problem into a large number of independent calculations.

### 2014-04-12 ThesisJP/Introduction/introduction.tex
Started on comments from Ines on the introduction section.

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision this is more than the worlds most powerful supercomputer could provide in 1998, and that was a room full of computers.

This immense rise in computational power and improved programmability are currently changing how ultrasound imaging systems are designed. Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software. And algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, becomes realizable. It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.

The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second. Nevertheless, with modern GPUs, these levels of processing are finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

Main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}). However, two additional methods were also explored. The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}). Both share the property of being computationally complex, but on the other hand they consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.

@@

\subsection{Paper\,I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
The first paper describes in details how the Capon beamformer was mapped to GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements. This makes it somewhat easier to reach the goal of real-time processing.

The estimation of the spatial covariance matrix receives special attention in this study. In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex step. In this paper we show that estimation of the sample covariance matrix is actually the most complex part when spatial and temporal smoothing with common parameters are included. It is then shown how the arithmetic complexity of this estimation process can be reduced from cubic to square. Finally, an in-depth analysis of the arithmetic throughput on multiple platforms is given. Despite our effort, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the matrix equation solver and covariance estimator respectively. Still, the reported throughput of 1 Mpx/s on a high-end GPU is enough to provide real-time processing for common sonar scan sequences.

\subsection{Paper\,II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance. This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade. In \textbf{Paper\,I} arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU. For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.  \textbf{Paper\,II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.

\subsection{Paper\,III}
\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication.}\\\\
In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer. As mentioned, shift-invariant imaging is essential if the method is ever to be adopted. The paper was written based on observations of aliasing artifacts when imaging bright point scatterers \textit{in vitro} and in simulations. The point scatterers were observed to twinkle like stars in the sky. In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated. Oversampling on transmit is not an option if the imaging modality is used to image rapidly varying objects. Smooth presentation of probe motion is equally important since real-time user interaction is one of the key selling points of ultrasound imaging. The frame rate should therefore be higher than 10 frames per second.

Further it has been investigated how the Capon beamforming achieves its super resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector. Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves from frame to frame and the beam density is insufficient. With a narrowband and farfield model it is shown that the Capon beamformer suffers from beam-to-beam gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.  To obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required. For broadband nearfield simulations the same gain variations are observed. Finally we show that the gain variation can be reduced, hence improved shift-invariance, by oversampling on receive using phase rotation steering. This is achieved without affecting the acquisition frame rate and with a minor increase in computational complexity. The method is successfully applied on simulations and \textit{in vitro} phantom data.

\subsection{Paper\,IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.

\subsection{Paper\,V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

\section{Main contributions}
The main contributions of this thesis are:
\begin{enumerate}
\item a GPU implementation of the Capon beamformer.
\item a GPU implementation of the beamspace Capon beamformer.
\item that real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
\item the first investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
\item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
\item a method for improved shift-invariance of the Capon beamformer when applied to medical ultrasound imaging.
\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
\item a GPU implementation of this adaptive volume rendering method.
\item a GPU implementation of simple Paint-like simulation tool for rapid visualization ultrasound pressure fields.
\end{enumerate}
In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.

@@

Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the maximum throughput of the target GPU. This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended. It might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed. This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU. An interesting observation can be found in Fig.\,10 of \textbf{Paper\,I}. Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}. This clearly shows the benefit of modern GPU computing.  Solving was performed by a batched Gauss Jordan solver implemented by Nvidia. In our work we focused on the covariance estimation step, and did not try to beat Nvidia at home. Nevertheless, with the measured throughput in mind, a natural next step would be to analyze the current solver, and figure out if any further optimization is possible. Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might also help to improve the solver's throughput.

>>>>

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision, this is more than the world's most powerful supercomputer could provide in 1998, and that was a room full of computers.

This immense rise in computational power and improved programmability are currently changing how ultrasound imaging systems are designed. Algorithms which previously had to be implemented in hardware for performance reasons can now be implemented in software. 
%Algorithms, either too complicated or costly to implement in hardware, and where a software implementation was thought to be too computationally heavy for real-time use, become realizable. 
It is clear that high-performance programmable processors, like the GPU, already have and will continue to make future ultrasound imaging systems more flexible, cheaper to produce, and equipped with even more cutting-edge processing.

The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of hundreds of GFLOPS is required. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays and an increasing number of samples per second. Nevertheless, with modern GPUs, these levels of processing are finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is supposed to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

The main focus has been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging [\textbf{Paper\,I} and \textbf{II}]. However, two additional methods have also been explored. The first one being adaptive visualization of cardiac ultrasound volumes [\textbf{Paper\,IV}], and the latter being simulation of dense ultrasound pressure fields [\textbf{Paper\,V}]. Both share the property of being computationally complex, but on the other hand they consist of many independent computations, which makes them perfectly suited for parallel GPU acceleration.

@@

\subsection{Paper\,I}
\textbf{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}\\
J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng\\
{\it IEEE Transactions on Oceanic Engineering, accepted for publication.}\\\\
The first paper describes in detail how the Capon beamformer is mapped to the GPU architecture. Even though the paper is written within the field of active \nom{sonar}{SOund Navigation And Ranging (usually under water)} imaging, the depicted implementation is applicable to a vast range of active imaging systems. A similar discussion for cardiac ultrasound imaging can be found in \textbf{Paper\,VI}.  Active sonar imaging typically differs from medical ultrasound imaging by a lower real-time requirement and fewer array elements. This makes it somewhat easier to reach the goal of real-time processing.

The estimation of the spatial covariance matrix receives a special attention in this study. In previous literature on Capon beamforming the matrix inversion has always been regarded as the most computationally complex step. In this paper we show that estimation of the sample covariance matrix is the most complex part when spatial and temporal smoothing of the estimate are included. It is then shown how the arithmetic complexity of this estimation procedure can be reduced from cubic to square. Finally, an in-depth analysis of the arithmetic throughput on multiple platforms is given. Despite our effort, the number of effective FLOPS is only 4 \% and 14 \% of the theoretical throughput of the target GPU for the matrix equation solver and covariance estimator respectively. Still, the reported throughput of 1 Mpx/s on a high-end GPU is enough to provide real-time processing for common sonar scan sequences.

\subsection{Paper\,II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance. This is something which has been suggested as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade. In \textbf{Paper\,I} arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging has typically 64 or more elements. In \textbf{Paper\,VI}, which is summarized in \textbf{Paper\,II}, it is shown that our implementation from  \textbf{Paper\,I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

In this paper, to reduce the matrix size, we take the advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU. For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.  \textbf{Paper\,II} and \textbf{VI} also present for the first time videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.

\subsection{Paper\,III}
\textbf{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}\\
\textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, accepted for publication.}\\\\
In the third paper we study the shift-invariant property of an imaging system based on the Capon beamformer. As mentioned, shift-invariant imaging is essential if the method is ever to be adopted. The paper is based on an observation of aliasing artifacts when the Capon beamformer was used to image bright point scatterers \textit{in vitro} and in simulations. The intensity of the point scatterers were observed to oscillate as they or the probe moved. In earlier work on Capon beamforming for medical ultrasound imaging the real-time nature of ultrasound imaging has often been ignored, and a large degree of oversampling on transmit might have been applied without being clearly stated. Oversampling on transmit will reduce the acquisition frame rate, and is therefore not an option if the imaging system is used to image rapidly varying objects. Smooth presentation of probe motion is equally important since real-time user interaction is one of the key selling points of ultrasound imaging.

Further, it has been investigated how the Capon beamforming achieves its super-resolution, and how this makes it highly sensitive to differences between the assumed steering vector and the signal propagation vector. Earlier work, focusing on single images, has typically referred to this effect as speed-of-sound errors, however, it also comes naturally into play when objects moves across an imaging sector which has insufficient lateral sampling. With a narrowband and farfield model it is shown that the Capon beamformer suffers from lateral gain variations as large as 27 dB for parameters previously used to obtain high resolution medical ultrasound images.  To obtain the same lateral gain variations as with \nom{DAS}{Delay-and-sum beamforming} beamforming, 25 times oversampling is shown to be required. For broadband nearfield simulations the same gain variations are observed. Finally we show that the gain variation can be reduced, hence improved shift-invariance, by oversampling on receive using phase rotation steering. This is achieved without affecting the acquisition frame rate and with a minor increase in computational complexity. The method is successfully applied on simulations and \textit{in vitro} phantom data.

\subsection{Paper\,IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
The fourth paper proposes an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood, as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.

\subsection{Paper\,V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646.}\\\\
The fifth paper presents an interactive simulation tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contribution in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox.} version. Second, the Paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

\section{Main contributions}
The main contributions of this thesis are:
\begin{enumerate}
\item a GPU implementation of the Capon beamformer.
\item a GPU implementation of the beamspace Capon beamformer.
\item a real-time beamspace Capon beamforming is achieved for cardiac ultrasound imaging.
\item a first-time investigation of Capon beamforming applied on multiple frames in medical ultrasound imaging (simulated, \textit{in vitro}, and \textit{in vivo}).
\item an investigation of shift-invariance when the Capon beamformer is applied on consecutive frames.
\item a method for improved shift-invariance of the Capon beamformer when applied to medical ultrasound imaging.
\item a method for reduced blood-pool noise in volume renderings of cardiac ultrasound volumes.
\item a GPU implementation of this adaptive volume rendering method.
\item a GPU implementation of simple Paint-like simulation tool for rapid visualization ultrasound pressure fields.
\end{enumerate}
In addition, the thesis provides several discussions on how to utilize the GPU to accelerate advanced ultrasound imaging algorithms.

@@

Even with all our effort in \textbf{Paper\,I} we were only able to achieve a fraction of the maximum throughput of the target GPU. This result shows how hard it is to reach the theoretical level of throughput, and that how close one can get is highly algorithm depended. There might be ways to improve on these numbers, but both the solver and covariance estimation step consist of several points where fine granular synchronization and serialized instructions are needed. This will restrict even the most fine tuned implementation from reaching the maximum throughput of the GPU. An interesting observation can be found in Fig.\,10 of \textbf{Paper\,I}. Here we see how the achieved throughput of the covariance estimation kernel is more than three times higher than the theoretical throughput of a modern, high-end, \nom{CPU}{Central processing unit}. This clearly shows the benefit of modern GPU computing.  Solving was performed by a batched Gauss Jordan solver implemented by Nvidia. In our work we focused on the covariance estimation step, and did not try to beat Nvidia at home. Nevertheless, with the measured throughput in mind, a natural next step would be to analyze the current solver, and figure out if any further optimization is possible. Since the sample covariance matrix is hermitian, a solver based on Cholesky decomposition might also help to improve the solver's throughput.
